---
title: 'Case Study 8: Diagnosing problems'
author: "Richard Wilkinson"
output:
  pdf_document:
    highlight: default
    toc: no
  html_document:
    theme: default
    toc: yes
---

Lets look at some data from a Canadian survey on wages in Ontario. The data contains information on

- wages: hourly rate
- education: number of years of schooling
- age in years
- sex
- language spoken

of 7425 individuals.

```{r, warning=FALSE}
library(car)
data(SLID)
str(SLID)
SLID[1:10,]
```

There are a number of missing data points. Lets remove these from the dataset and only work with individuals for whom we have complete information.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
WageData <- filter(SLID, !is.na(wages), !is.na(education), !is.na(language))
scatterplotMatrix(WageData[,c('wages','sex', 'age', 'education')])
```

Lets start by fitting a basic model and examining some diagnostic plots.
```{r}
fit <- lm(wages~ sex + age + education, data = WageData)
qqPlot(fit)
hist(rstudent(fit))
```

The QQ-plot and the histogram of the residuals suggests a positive skew (as did the original histogram of the wage data in the scatterplot matrix). We can reduce positive skew by moving the response variable down the ladder of powers.

```{r}
fit2 <- lm(log(wages) ~ sex + age + education, data = WageData)
qqPlot(fit2)
```

You could try other powers such as sqrt(wage), but log(wage) seems to have produced reasonable results. 
Note that you can highlight the 5 most outlying data points in QQ-plots by using the command
```{r, eval=FALSE}
qqPlot(fit2, id.n=5)
```

Lets now check the residual plots:
```{r}
residualPlots(fit2, tests=FALSE)
crPlots(fit2, terms=~age+education)
```

The command 
```{r, eval=FALSE}
crPlots(fit2) 
```

will also work, but gives a third plot for the categorical variable sex which is not useful. So I've removed it by specifying the terms to plot with the command crPlots(fit2, terms=~age+education).

This shows some non-linearity, particularly in how the response depends upon age. We could try transforming age down the ladder of powers, but the  relationship looks like it could be non-monotone, in which case we will need to fit a quadratic model. Similarly, the dependence upon education looks  non-linear. We can try transforming this up the ladder of powers (see Tukey and Mosteller's bulging rule).



```{r, message=FALSE}
fit3 <- lm(log(wages) ~ sex + age + I(age^2) + I(education^2), WageData) 
residualPlots(fit3, tests=FALSE)
crPlots(fit3, terms=~.-sex)
```

These now looks pretty much perfect.
